### 2018-06-29 report.tex
Chapter 1 finished

@@

%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
\setcounter{tocdepth}{1}
\tableofcontents % Indholdsfortegnelse
%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Protein research can be approached from different angles. This thesis exploited the combination of mass spectrometry \ac{MS} and proteomics workflows (see \ref{chap:mass_spec}) to the qualitative and quantitative characterization of protein samples.

@@

An overview over the \ac{MS} and downstream computational analysis of its data is presented in \ref{chap:mass_spec}. The pipeline development and its benchmark are explained in chapters 3 and 4, while the modelling problem is introduced in chapter 5. Finally, a conclusion of the work is given in chapter 6.

\chapter{Mass spectrometry and shotgun proteomics}
\label{chap:mass_spec}

@@

A summary of the bottom-up approach MS analytical pipeline follows.

@@

\section{The mass spectrometer}
\label{sec:thee_mass_spectrometer}

The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector SEE FIGURE.

@@

On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values. They constitute the isotopic envelope of the ion SEE FIGURE, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms. Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da. Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise. Such event can be problematic if it entails that the 1 \ce{^{13}_{}C} peak is confused for the completely \ce{^{12}_{}C} peak.

@@

\ac{HPLC} methods work by loading the peptide mix in a column containing a stationary and a solid phase. These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases. The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times. Therefore, the input to the machine will consist of a simplified mix of peptides at a time. Notably, the same peptide could elute in several contiguous fractions.

@@

Given the stochastic nature of the cleavage and spectra recording process, the resulting spectra exhibit variability manifested in missing peaks or spurious ones. Furthermore, the PSM process against a sufficiently big database can, at random, return wrong matches. This translates to the generation of multiple matches, of which one, if any, will be correct. Therefore, the lists of matches need to be somehow ranked. The issue is addressed by search engines through the deployment of statistical models that provide scoring systems measuring the goodness of the match. Assuming the correct protein is present in the database, a good scoring system should give the matched peptide which comes from the right protein the best score. Under this circumstances, enough evidence for the presence of individual proteins can be collected.

@@

\item Peptide inference: infer the peptides present in the sample.
\item Protein inferece: based on the inferred peptides, infer what proteins generated them. This is not trivial as the peptide coming from the best PSM could map to more than one protein.
\end{enumerate}

Both are taken care of by the search engine. Multiple search engines exist that implement different matching and scoring algorithms. The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda. Notably, the results of each individual search engine can be combined to gather their strenghts, at the expense of an increased computational cost and time REF THIS!!

%\subsection{X!Tandem}
%
%X!Tandem was one of the first open source search engines in mass spectrometry. It produces a score based on the dot product between the theoretical and the experimental tandem mass spectra CITE Overview of Tandem Mass Spectrometry (MS/MS) Database Search Algorithms. The scores assigned to wrong matches are assumed to follow a hypergeometric distribution, allowing the program to extrapolate $E(s)$ (expected number of wrong matches at a given score) for any score.

\section{Results validation}
\label{sec:validation}

The most commonly used method to compute the FDR of a list of matches is the target-decoy search. Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database. This decoy database is generated by reversing or in general applying a randomization technique upon the sequences present in the original database (target). This guarantess that while the sequences in the decoy database won't overlap with those in the target, their basic properties of the decoy (size, composition, etc) will remain identical to the target.

All matches to the decoy are by definition wrong. Since database properties 
are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be used to estimate the number of false identifications in the  target results. it is as likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5, thus the number of matches in the decoy provides an accurate estimate of the number of random matches, or false positives, against the target database ($n_{fp}$). Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.

$$FDR = \frac{n_{fp}}{n_{tp} + n_{tp}}$$

The minimal FDR at which a given PSM is considered a positive match constitutes the PSM\textquotesingle s q-value

$$ q(PSM_i) = \min FDR \forall PSM_i \in Positives$$

Finally, a posterior error probability (PEP) can be defined as the probability of the match being random ($P(s_i | T_i = 0)$). This can only be done if a statistical model describing the distribution of scores for correct and wrong matches is fit to the dataset.

\section{Protein quantification}

CONTINUE maybe add reason for the project.

\chapter{Materials and Methods}

\chapter{Results and Discussion}

\chapter{Conclusion}

@@

\section{Overview of proteomics analysis software}

\subsection{MaxQuant}

One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}. It can be downloaded for free and is developed by the Max Planck institute in Germany. It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.

MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline. A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.

The main processing steps incorporated in MaxQuant consist of:

\begin{itemize}

\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.

\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.

\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.

\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.

\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.

\item Supports label-based and label-free quantification.

\item Its output can be passed to the Perseus software for data visualization. %CITE

\end{itemize}

However, MaxQuant also suffers from some problems:

\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}. The processing steps cannot easily be fine tuned or exchanged with other pipelines. For example, only the Andromeda search engine is supported. This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.

\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.

\item While it is supported in Linux through Mono, the user experience is best in Windows. As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.

\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.

\subsection{Compomics group}

The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.

\begin{itemize}

\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.

\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.

\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide. Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.

\end{itemize}

SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}. Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}. They easily enable the development of analytical pipelines finely fitted to each use case.

moFF is less documented and integrated in the workflow. Furthermore, it does not offer a GUI yet. The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.

The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool. Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.

\subsection{Other suites}

Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available. They are not open, requiring a license to function. Furthermore, they are GUI oriented and make use of different data formats thus, making exchange across pipelines extremely difficult. The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.

\section{Label-free quantification}

\subsection{MaxLFQ}

The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}. It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.

\subsubsection{Fractionated XIC aggreagation}

First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions. MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity 
 between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation).

The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
%  this commented blank line prevents start of a new paragraph
\begin{equation}\label{eq:peptide_intensity}
I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
\end{equation}

The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
%  this commented blank line prevents start of a new paragraph
\begin{equation}\label{eq:sum_norm_intensity}
H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
\end{equation}
%  this commented blank line prevents start of a new paragraph
\begin{equation}
H(N) = \sum\limits_{p} H_p(N) 
\end{equation}

where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.

The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{xic_normalization}
\caption{Taken from \cite{Cox2014}}
\label{figure:xic_normalization}
\end{figure}

\subsubsection{Protein intensity inference}

Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred. This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).

\begin{enumerate}
\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.

\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b. The median is selected as a summarising statistic to protect from outliers. A minimal number of non-zero intensity peptides are required for the median to be valid, usually 2. Otherwise, the protein ratio for the pair of samples is set to 0.

\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.

\begin{equation}\label{eq:least_squares}
\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
\end{equation}

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{protein_quant_inference}
\caption{Taken from \cite{Cox2014}}
\label{figure:protein_quant_inference}
\end{figure}

>>>>

\DeclareAcronym{FDR}{
  short = FDR ,
  long  = False Discovery Rate,
  class = abbrev
}

\DeclareAcronym{RT}{
  short = RT ,
  long  = Retention time,
  class = abbrev
}

\DeclareAcronym{MBR}{
  short = MBR ,
  long  = Match Between Runs,
  class = abbrev
}

\DeclareAcronym{XIC}{
  short = XIC ,
  long  = Extracted Ion Current,
  class = abbrev
}

\DeclareAcronym{SC}{
  short = SC ,
  long  = Spectral Counting,
  class = abbrev
}

\DeclareAcronym{TPP}{
  short = TPP ,
  long  = Trans Proteomic Pipeline,
  class = abbrev
}

@@

%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
\setcounter{tocdepth}{1}
\tableofcontents % Indholdsfortegnelse
%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

@@

Protein research can be approached from different angles. This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.

@@

\section{Structure of the Thesis}

An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}. The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}. Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.

\chapter{Mass spectrometry and shotgun proteomics overview}
\label{chap:mass_spec}

@@

A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:

\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:mass_spectrometry_workflows}.

\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.

@@

\section{The mass spectrometer}
\label{sec:the_mass_spectrometer}

The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).

\begin{figure}[!h]
\includegraphics[width=\textwidth]{mass_spectrometer}
\caption{Schematic view of a mass spectrometer. Taken from}
%CITE THIS https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation
\label{fig:mass_spectrometer}
\end{figure}

@@

On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values. They constitute the isotopic envelope of the ion SEE FIGURE, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms. Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da. Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise. Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.

@@

\ac{HPLC} methods work by loading the peptide mix in a column containing a stationary and a solid phase. These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases. The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer. Therefore, the input to the machine will consist of a simplified mix of peptides at a time. Notably, the same peptide could elute in several contiguous fractions.

@@

Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones. Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database. This translates to the generation of multiple matches, of which one, if any, will be correct. Therefore, the lists of matches need to be somehow ranked. The issue is addressed by search engines through the deployment of statistical models that provide scoring systems measuring the goodness of the match. Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide. Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.

Multiple search engines exist that implement different matching and scoring algorithms. The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda CITE THE ENGINES. Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time REF THIS!!

\section{Validation and quality control}
\label{sec:validation}

The most commonly used method to compute the FDR of a list of matches is the target-decoy search. Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database. This decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target). This guarantess that while the sequences in the decoy database won't overlap with those in the target, their basic properties of the decoy (size, composition, etc) will remain identical to the target.

All matches to the decoy are by definition wrong. Since database properties 
are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the equivalent list of target results. This is because the existence of shared properties entails that random matches are equally likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5. Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.

\begin{equation}\label{eq:fdr}
FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
\end{equation}

Equation \ref{eq:fdr} can be used to select the number of positives (the denominator of the fraction) that make the FDR equal to a predefined value, frequently 0.01 or 1 \%.

%The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value
%
%\begin{equation}\label{eq:qvalue}
%q(PSM_i) = \min FDR \forall PSM_i \in Positive
%\end{equation}

\section{Peptide and protein inference}
\label{sec:inference}

@@

\item \textbf{Peptide inference}: infer the peptides present in the sample.
\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them. This is not trivial as peptides are degenerate and could map to more than one protein.
\end{enumerate}

Peptide inference is performed during the PSM process. The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms. The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one. Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.

\section{Protein quantification}
\label{sec:quantification}

The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample. However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone. For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.

Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free). These are explained in subsection \ref{subsec:labelling}. If the label-free approach is employed, more distinctions can be made based on:

\begin{itemize}
\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}

\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based.
\end{itemize}

\subsection{Label-based and label-free approaches}
\label{subsec:labelling}

Two paradigms exist in protein quantification: label-based and label-free. In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label. All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability CITE PAGE 237 COMPUTATIONAL METHODS The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified CITE MAXLFQ. Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) CITE or Isotope-Coded Affinity Tags (ICAT) CITE, differ in the nature of the label and the way it is introduced. Remarkably, peptide labeling costs can be quite high. This, together with the limited amount of samples that can be compared makes a case for label-free quantification.

In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs. In order to account for inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing can be carried out for pair-wise transfer of peptide identities from runs where identification was successful to those were it wasn't due to the randomness in the identification process. The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF.

\subsection{SC and XIC based quantification}
\label{subsec:scvsxic}

Quantification can be spectral counting or \ac{XIC} based.
Spectrum counting based quantification is the simplest quantification method in proteomics. It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra. As a result, peptides from abundant proteins are more likely to be
identified and in more spectra. Two approaches were hence followed: count the number of peptides identified for a given protein – like in the emPAI index method – or count the number of spectra ascribed to a protein as in the NSAF method CITE COMPOMICS 4.1. These methods have the advantage that they are very simple to implement and don't require any further data processing.

\ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance CITE. A wide range of algorithms are available to process these data and output estimates of protein abundance. All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements. They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.

Although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms. These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra. CITE MAXLFQ For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.

\subsection{\ac{XIC}-peptide-based models for label-free quantification}
\label{subsec:peptide_model}

The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample. However, the data interpretation requires the transfer of these peptide-level data into the protein level. This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.

As stated in  CITE SUMMARY MSQROB: \textit{Peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates. However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein. This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.

It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible. This translates into improved results when compared to the other aforementioned methods CITE MSQROB SUMMARY.

%CITE MAXQUANT
%
% The MaxLFQ algorithm, provided within the MaxQuant suite, is an \ac{MS1}, summarization method performing two optimisation steps to

\chapter{Assembly of an open-source Linux-based label-free quantification proteomics pipeline}
\label{chap:pipeline}

\section{Introduction}

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented modelling frameworks and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) or openMS  CITE ALL are nevertheless available for the community. MaxQuant, a free but closed-source proteomics analysis suite CITE MAXQUANT, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.

However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform. As a GUI would not be required to run the pipeline, it would also be scalable to big datasets. Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.

In this chapter, a pipeline making use of the set of tools published by the Compomics  and StatOmics groups CITE THEM, which fulfill all the foregoing requirements, will be presented.

\section{Materials and Methods}

\subsection{Data generation}

\subsection{Data loading}

\subsection{Database search}

\subsection{Quality control and validation}

\subsection{Data refinement}

\subsection{Quantification}

\section{Results and Discussion}

\section{Conclusion}

\chapter{Pipeline benchmarking on a real biological setting}
\label{chap:benchmark}

\chapter{Bayesian approach to protein quantification}
\label{chap:model}

@@

\chapter{Conclusion of the Thesis}
\label{chap:conclusion}

%\section{Overview of proteomics analysis software}
%
%\subsection{MaxQuant}
%
%One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}. It can be downloaded for free and is developed by the Max Planck institute in Germany. It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.
%
%MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline. A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.
%
%The main processing steps incorporated in MaxQuant consist of:
%
%\begin{itemize}
%
%\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.
%
%\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.
%
%\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.
%
%\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.
%
%\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.
%
%\item Supports label-based and label-free quantification.
%
%\item Its output can be passed to the Perseus software for data visualization. %CITE
%
%\end{itemize}
%
%However, MaxQuant also suffers from some problems:
%
%\begin{enumerate}
%
%\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}. The processing steps cannot easily be fine tuned or exchanged with other pipelines. For example, only the Andromeda search engine is supported. This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.
%
%\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.
%
%\item While it is supported in Linux through Mono, the user experience is best in Windows. As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.
%
%\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.
%
%\end{enumerate}
%
%\subsection{Compomics group}
%
%The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.
%
%\begin{itemize}
%
%\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.
%
%\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.
%
%\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide. Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.
%
%\end{itemize}
%
%SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}. Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}. They easily enable the development of analytical pipelines finely fitted to each use case.
%
%moFF is less documented and integrated in the workflow. Furthermore, it does not offer a GUI yet. The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.
%
%The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool. Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.
%
%
%\subsection{Other suites}
%
%Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available. They are not open, requiring a license to function. Furthermore, they are GUI oriented and make use of different data formats thus, making exchange across pipelines extremely difficult. The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.
%
%\section{Label-free quantification}
%
%\subsection{MaxLFQ}
%
%The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}. It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.
%
%\subsubsection{Fractionated XIC aggreagation}
%
%First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions. MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity 
% between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation). 
%
%The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}\label{eq:peptide_intensity}
%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
%\end{equation}
%
%
%The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}\label{eq:sum_norm_intensity}
%H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
%\end{equation}
%%  this commented blank line prevents start of a new paragraph
%\begin{equation}
%H(N) = \sum\limits_{p} H_p(N) 
%\end{equation}
%
%
%where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.
%
%The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.
%
%\begin{figure}[!h]
%\centering
%\includegraphics[width=0.8\textwidth]{xic_normalization}
%\caption{Taken from \cite{Cox2014}}
%\label{figure:xic_normalization}
%\end{figure}
%
%
%\subsubsection{Protein intensity inference}
%
%Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred. This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).
%
%\begin{enumerate}
%\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.
%
%\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b. The median is selected as a summarising statistic to protect from outliers. A minimal number of non-zero intensity peptides are required for the median to be valid, usually 2. Otherwise, the protein ratio for the pair of samples is set to 0.
%
%\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.
%
%\end{enumerate}
%
%\begin{equation}\label{eq:least_squares}
